---
title: "Base R Graphics"
output: 
  learnr::tutorial:
    css: css/styles.scss
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: |
  Provides a basic introduction to R Graphics
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)

gradethis_setup(exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE)

mtcars_4 <- subset(mtcars, cyl == "4")
mtcars_6 <- subset(mtcars, cyl == "6")
mtcars_8 <- subset(mtcars, cyl == "8")
```

## Introduction {data-progressive="FALSE"}

This tutorial provides an introduction on how to create different graphics in R. For this tutorial, we will focus on plotting different components from the `mtcars` data set.


::: contents
### Contents

1.  Basic

2.  Grouping

3.  Tweaking
:::

### Notation

Through out this tutorial, we use certain notations for different components in R. To begin, when something is in a gray block, `_`, this indicates that R code is being used. When I am talking about an R Object, it will be displayed as a word. For example, we will be using the R object `mtcars`. When I am talking about an R function, it will be displayed as a word followed by an open and close parentheses. For example, we will use the mean function denoted as `mean()` (read this as "mean function"). When I am talking about an R argument for a function, it will be displayed as a word following by an equal sign. For example, we will use the data argument denoted as `data=` (read this as "data argument"). When I am referencing an R package, I will use `::` (two colons) after the name. For example, an alternative to R graphics is using the `ggplot2::` (read this as "ggplot2 package") Lastly, if I am displaying R code for your reference or to run, it will be displayed on its own line. There are many components in R, and my hope is that this will help you understand what components am I talking about.

## Basic Graphics

Here we will use the built-in R functions to create different graphics. The main function that you will use is the `plot()`. It contains much of the functionality to create many different plots in R. Additionally, it works well for different classes of R objects. It will provide many important plots that you will need for a certain statistical analysis.

### Scatter Plot

Let's first create a scatter plot for one variable using the `mpg` variable from the `mtcars` data set. This is done using the `plot()` and setting the first argument `x=` to the vector.

```{r base_sp_1, exercise = TRUE}

```

```{r base_sp_1-solution}
plot(mtcars$mpg)
```

```{r base_sp_1-check}
grade_this_code()
```

Notice that the x-axis is the index (which is not informative) and the y-axis is the `mpg` values.

Let's connect the points with a line. This is done by setting the `type=` to `"l"`.

```{r base_sp_2, exercise = TRUE}

```

```{r base_sp_2-solution}
plot(mtcars$mpg, type = "l")
```

```{r base_sp_2-check}
grade_this_code()
```


Let's add the points back to the plot and keep the lines. What we are going to do is first create the scatter plot as we did before, but we will also use the `lines()` to add the lines. The `lines()` needs the `x=` which is a vector of points (`mpg`). The two lines of code must run together.

```{r base_sp_3, exercise = TRUE}

```

```{r base_sp_3-solution}
plot(mtcars$mpg)
lines(mtcars$mpg)
```

```{r base_sp_3-check}
grade_this_code()
```

Now, let's create a more realistic scatter plot with 2 variables. This is done by specifying the `y=` with another variable in addition to the `x=` in the `plot()`. Plot a scatter plot between `mpg` and `disp`.

```{r base_sp_4, exercise = TRUE}

```

```{r base_sp_4-solution}
plot(mtcars$mpg,mtcars$disp)
```

```{r base_sp_4-check}
grade_this_code()
```

Now, let's change the the axis labels and plot title. This is done by using the arguments `main=`, `xlab=`, and `ylab`. The `main=` changes the title of the plot.

### Histogram

To create a histogram, use the `hist()`. The `hist()` only needs `x=` which is numerical vector. Create a histogram with the `mpg` variable.

```{r base_hist_1, exercise = TRUE}

```

```{r base_hist_1-solution}
hist(mtcars$mpg)
```

```{r base_hist_1-check}
grade_this_code()
```

If you want to change the number of breaks in the histogram, use the `breaks=`. Create a new histogram of the `mpg` variable with ten breaks.

```{r base_hist_2, exercise = TRUE}

```

```{r base_hist_2-solution}
hist(mtcars$mpg, breaks = 10)
```

```{r base_hist_2-check}
grade_this_code()
```

The above histograms provide frequencies instead of relative frequencies. If you want relative frequencies, use the `freq=` and set it equal to `FALSE` in the `hist()`.

```{r base_hist_3, exercise = TRUE}

```

```{r base_hist_3-solution}
hist(mtcars$mpg, freq = FALSE)
```

```{r base_hist_3-check}
grade_this_code()
```

### Density Plot

A density plot can be used instead of a histogram. This is done by using the `density()` to create an object containing the information to create density function. Then, use the `plot()` to display the plot. The only argument the `density()` needs is the `x=` which is the data to be used. Create a density plot the `mpg` variable.

```{r base_dense_1, exercise = TRUE}

```

```{r base_dense_1-solution}
plot(density(mtcars$mpg))
```

```{r base_dense_1-check}
grade_this_code()
```

Now, if we want to overlay the density function over a histogram, use the `lines()` with the output from the `density()` as its main input. First create the histogram using the `hist()` and setting the `freq=` to `FALSE`. Then use the `lines()` to overlay the density. Make sure to run both lines together.

```{r base_dense_2, exercise = TRUE}

```

```{r base_dense_2-solution}
hist(mtcars$mpg, freq = FALSE)
lines(density(mtcars$mpg))
```

```{r base_dense_2-check}
grade_this_code()
```

### Box Plots

A commonly used plot to display relevant statistics is the box plot. To create a box plot use the `boxplot()`. The function only needs the `x=` which specifies the data to create the box plot. Use the box plot function to create a box plot on for the variable `mpg`.

```{r base_box_1, exercise = TRUE}

```

```{r base_box_1-solution}
boxplot(mtcars$mpg)
```

```{r base_box_1-check}
grade_this_code()
```

If you want to make the box plot horizontal, use `horizontal=` and set it equal to `TRUE`.

```{r base_box_2, exercise = TRUE}

```

```{r base_box_2-solution}
boxplot(mtcars$mpg, horizontal = TRUE)
```

```{r base_box_2-check}
grade_this_code()
```

### Bar Chart

A histogram shows you the frequency for a continuous variable. A bar chart will show you the frequency of a categorical or discrete variable. To create a bar chart, use the `barplot()`. The main argument it needs is the `height=` which needs to an object from the `table()`. Create a bar chart for the `cyl` variable.

```{r base_bar_1, exercise = TRUE}

```

```{r base_bar_1-solution}
barplot(table(mtcars$cyl))
```

```{r base_bar_1-check}
grade_this_code()
```

### Pie Chart

While I do not recommend using a pie chart, R is capable of creating one using the `pie()`. It only needs the `x=` which is a vector numerical quantities. This could be the output from the `table()`. Create a pie chart with the `cyl` variable.

```{r base_pie_1, exercise = TRUE}

```

```{r base_pie_1-solution}
pie(table(mtcars$cyl))
```

```{r base_pie_1-check}
grade_this_code()
```

## Grouping

Similar to obtaining statistics for certain groups, plots can be grouped to reveal certain trends. We will look at a couple of methods to visualize different groups.

### One Variable Grouping

Two ways to display groups is by using color coding or panels. I will show you what I think is the best way to group variables. There may be better ways to do this, such as using the ggplot2 package. Before we begin, create three new R objects that are a subset of the `mtcars` data set into 3 different data sets with for the three different values of the `cyl` variable: "4", "6", and "8". use the `subset()` to create the different data sets. Name the new R objects `mtcars_4`, `mtcars_6`, and `mtcars_8`, respectively.


```{r base_one_null_color, exercise = TRUE}

```

```{r base_one_null_color-solution}
mtcars_4 <- subset(mtcars, cyl == 4)
mtcars_6 <- subset(mtcars, cyl == 6)
mtcars_8 <- subset(mtcars, cyl == 8)
```

```{r base_one_null_color-check}
grade_this_code()
```

#### Scatter Plot

To create different colors points for their respective label associated `cyl` variable. First create a base scatter plot using the `plot()` to set up the plot. Then one by one, overlay a set of new points on the base plot using the `points()`. The first two arguments should be the vectors of data from their respective R object subset. Also, use the `col=` to change the color of the points. The `col=` takes either a string or a number.

```{r base_one_scatter_color, exercise = TRUE}

```

```{r base_one_scatter_color-solution}
plot(mtcars$mpg, mtcars$disp)
points(mtcars_4$mpg, mtcars_4$disp, col = "red")
points(mtcars_6$mpg, mtcars_6$disp, col = "blue")
points(mtcars_8$mpg, mtcars_8$disp, col = "green")
```

```{r base_one_scatter_color-check}
grade_this_code()
```

#### Histogram

Now, it us more difficult to overlay histograms on a plot to different colors. Therefore, a panel approach may be more beneficial. This can be done by setting up R to plot a grid of plots. To do this, use the `par()` to tell R how to set up the grid. Then use the `mfrow=`, which is a vector of length two, to set up a grid. The `mfrow=` usually has an input of `c(ROWS,COLS)` which states the number of rows and the number of columns. Once this is done, the next plots you create will be used to populate the grid.

```{r base_one_hist_panel, exercise = TRUE}

```

```{r base_one_hist_panel-solution}
par(mfrow=c(1,3))
hist(mtcars_4$mpg)
hist(mtcars_6$mpg)
hist(mtcars_8$mpg)
```

```{r base_one_hist_panel-check}
grade_this_code()
```

Every time you use the `par()`, it will change how graphics are created in an R session. Therefore, all your plots will follow the new graphic parameters. You will need to reset it by typing `dev.off()`.

#### Bar Chart

To visualize two categorical variables, we can use a color-coded bar chart to compare the frequencies of the categories. This is simple to do with the `barplot()`. First, use the `table()` to create a cross-tabulation of the frequencies for two variables. Then use the `boxplot()` to visualize both variables. Then use `legend=` to create a label when the bar chart is color-coded. Additionally, use the `beside=` argument to change how the plot looks. Use the code below to compare the variables `cyl` and `am` variable.

```         
barplot(table(mtcars$cyl, mtcars$am), beside = TRUE, legend = rownames(table(mtcars$cyl, mtcars$am)))
```

```{r base_bar, exercise = TRUE}

```

```{r base_bar-solution}
barplot(table(mtcars$cyl, mtcars$am), beside = TRUE, legend = rownames(table(mtcars$cyl, mtcars$am)))
```

```{r base_bar-check}
grade_this_code()
```

Notice that I use the `rownames()` to label the legend.

## Tweaking

### Labels

The main tweaking of plots I will talk about is changing the the axis label and titles. For the most part, each function allows you to use the `main=`, `xlab=`, and `ylab=`. The `main=` allows you to change the title. The `xlab=` and `ylab=` allow you to change the labels for the x-axis and y-axis, respectively. Create a scatter plot for the variables `mpg` and `disp` and change the labels.

```{r base_label, exercise = TRUE}

```

```{r base_label-solution}
plot(mtcars$mpg, mtcars$disp, main = "MPG vs Displacement", xlab = "MPG", ylab = "Displacement")
```

```{r base_label-check}
grade_this_code()
```


## Quiz

```{r quiz, echo=FALSE}
quiz(caption = "",
  question("If I were to extract a variable from a data frame, what type of R object is it?",
    answer("list", message = "A list contains multiple R objects"),
    answer("matrix", message = sprintf("A matrix produces a $n \\times k$ R object")),
    answer("vector", correct = TRUE),
    answer("data frame", message = "A data frame contains multiple variables"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("What are methods to import data into R?",
    answer("Base R Code", correct = TRUE),
    answer("shiny app", message = "Shiny apps are used to create data dashboards"),
    answer("`readr::` code", correct = TRUE),
    answer("RStudio data import wizard.", correct = TRUE),
    incorrect = "Incorrect. You are at least on one more thing.",
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
    question("When testing the type of variables, we utilized a function similar to `is.xxx()`. What type of data was the output?",
    answer("Logical", correct = TRUE),
    answer("Character", message = "A character will be surrounded by quotes"),
    answer("Numeric", message = "A number will be produced instead"),
    answer("Raw", message = "Raw would look like byte code"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("What is the difference between a matrix and a data frame?",
    answer("A matrix requires the data to be the same type and data frame does not", correct = TRUE),
    answer("Each row of a data frame corresponds to an experimental unit and a matrix does not.", message = "The rows of a matrix can still correspond to different experimental units and columns correspond to characteristics"),
    answer("A data frame can be indexed with the `$` operator and a matrix cannot.", correct = TRUE),
    answer("A matrix must be a squared and a data frame does not."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("Different programming languages have their index starting value at 0?",
           answer("TRUE", correct = T),
           answer("FALSE"),
           allow_retry = TRUE),
  question("Choose the correct brackets to index and R object:",
           answer("()"),
           answer("{}"),
           answer("[]", correct = TRUE),
           allow_retry = TRUE,
           random_answer_order = TRUE
           ),
  question_text(
  "Write a reflection on what you had learned about R objects.",
  answer(NULL, correct = TRUE),
  trim = TRUE,
  rows = 4,
  correct = NULL,
  incorrect = NULL,
  allow_retry = TRUE)

  
)
```

## Submit Responses

**Generate a Hash if you need to submit your work for an assignment.**

```{r context="server"}
learnrhash::encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please click the button below to generate your hash and",
    "submit it using the corresponding tutorial assignment tab on your Learning Management System, if needed.",
    shiny::tags$br()
  ))
```

## Feedback

```{=html}
<iframe src="https://docs.google.com/forms/d/e/1FAIpQLSda_bV4VFGk1pgrevik7afgjAlWTA0g9kQE7ZR52S_UQiT3NA/viewform?embedded=true" width="640" height="742" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>
```
