---
title: "R Strategies"
output: 
  learnr::tutorial:
    css: css/styles.scss
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: |
  Provides general tips and tricks to 
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)

gradethis_setup(exercise.reveal_solution = FALSE)
knitr::opts_chunk$set(echo = FALSE)


```


## R Packages

As I stated before, R can be extended to do more things, such as create this tutorial. This is done by installing R packages. An R package can be thought of as extra software. This allows you to do more with R. To install an R package, you will need to use `install.packages("NAME_OF_PACKAGE")`. Once you install it, you do not need to install it again. To use the R package, use `library("NAME_OF_PACKAGE")`. This allows you to load the package in R. You will need to load the package every time you start R. For more information, please watch the video below.

```{=html}
<iframe src="https://player.vimeo.com/video/203516241?h=819422ce1e&color=428bca" width="640" height="400" frameborder="0" allow="autoplay; fullscreen; picture-in-picture" allowfullscreen></iframe>
<p><a href="https://vimeo.com/203516241">install Packages</a> from <a href="https://vimeo.com/rstudioinc">RStudio, Inc.</a> on <a href="https://vimeo.com">Vimeo</a>.</p>
```

## Diagnostics

Here is more practice in trying to understand some of the errors R may output. Type the following code
```
qnorm(2)
```

```{r rd_1, exercise = TRUE}

```

```{r rd_1-solution}
qnorm(2)
```


Here a `NaN` is produced. Notice how R puts a warning in the output. Here it is much more difficult to diagnose what the problem may be. My first advice is to look at the R documentation for the `qnorm()`. Now looking at the documentation for this particular function may not provide information on why it is wrong. The next best thing may be to go online. However, you may have read that this function provides a quantile; therefore, the value we input must be a number between 0 and 1. 


## Quiz

```{r quiz, echo=FALSE}
quiz(caption = "",
  question("Which function do you need to use to properly load an R package?",
    answer("install.packages()", message = "This will install an R package."),
    answer("load()", message = sprintf("This is used to load data, not a package.")),
    answer("library()", correct = TRUE),
    answer("require()", message = "While this will work, it is advised not to use this function."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("What is the code to add 5 to 4 squared?",
    answer("5 + 4**2", correct = TRUE),
    answer("5 * 4 && 4", message = "Shiny apps are used to create data dashboards"),
    answer("4^2 + 5", correct = TRUE),
    answer("4*4 + 5", correct = TRUE),
    incorrect = "Incorrect. There is at least one more thing.",
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
    question("Which operator is used to create an R object with data? ",
    answer("<-", correct = TRUE),
    answer("==", message = "This is the absolute equal operator."),
    answer("=", correct = TRUE),
    answer("<<", message = "This is not a known operator"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("You will need to install each R package everytime you boot up R.",
           answer("TRUE"),
           answer("FALSE", correct = T),
           allow_retry = TRUE),
  question("Which one is the pipe operator?",
    answer("|>", correct = TRUE),
    answer("%|%", message = "This is not a known R operator."),
    answer("%>%", correct = TRUE),
    answer("%%", message = "This is the modulo (remainder) operator."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("Do you need to specify the name of the argument for a function?",
           answer("TRUE"),
           answer("FALSE", correct = T),
           allow_retry = TRUE),
  question("Which function tells you if a data value is a Logical value?",
           answer("if.logical()"),
           answer("as.logical()"),
           answer("is.logical()", correct = TRUE),
           allow_retry = TRUE,
           random_answer_order = TRUE
           ),
  question_text(
  "Write a reflection on what you had learned about R Basics.",
  answer(NULL, correct = TRUE),
  trim = TRUE,
  rows = 4,
  try_again_button = "Edit Answer",
  correct = NULL,
  incorrect = NULL,
  allow_retry = TRUE)

  
)
```

## Submit Responses

**Generate a Hash if you need to submit your work for an assignment.**

```{r context="server"}
learnrhash::encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please click the button below to generate your hash and",
    "submit it using the corresponding tutorial assignment tab on your Learning Management System, if needed.",
    shiny::tags$br()
  ))
```

## Feedback

```{=html}
<iframe src="https://docs.google.com/forms/d/e/1FAIpQLSda_bV4VFGk1pgrevik7afgjAlWTA0g9kQE7ZR52S_UQiT3NA/viewform?embedded=true" width="640" height="742" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>
```
