---
title: "Basic Probability and Statistics"
output: 
  learnr::tutorial:
    css: css/styles.scss
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: |
  This tutorial provides code to conduct basic statistics.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)

gradethis_setup(exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```


## Introduction {data-progressive="FALSE"}

This tutorial provides an introduction on some of the basic statistics in R.

::: contents
### Contents

1. Probability

2. Statistics

:::

```{r intro, echo=FALSE}
question_text(
  "Make a prediction on how R calculates basic probability and descriptive statistics.",  answer(NULL, correct = TRUE),
  trim = TRUE,
  rows = 4,
  correct = NULL,
  incorrect = NULL,
  try_again_button = "Edit Answer",
  allow_retry = TRUE)

```

### Notation

Through out this tutorial, we use certain notations for different components in R. To begin, when something is in a gray block, `_`, this indicates that R code is being used. When I am talking about an R Object, it will be displayed as a word. For example, we will be using the R object `mtcars`. When I am talking about an R function, it will be displayed as a word followed by an open and close parentheses. For example, we will use the mean function denoted as `mean()` (read this as "mean function"). When I am talking about an R argument for a function, it will be displayed as a word following by an equal sign. For example, we will use the data argument denoted as `data=` (read this as "data argument"). When I am referencing an R package, I will use `::` (two colons) after the name. For example, an alternative to R graphics is using the `ggplot2::` (read this as "ggplot2 package") Lastly, if I am displaying R code for your reference or to run, it will be displayed on its own line. 

## Probability

We will discuss several R functions used for different probability distributions

### Common Distributions

1. **Uniform Distribution**: In a uniform distribution, all outcomes within a given range are equally likely. For example, rolling a fair six-sided die where each face has an equal probability of 1/6 (Discrete Version).

2. **Normal Distribution (Gaussian Distribution)**: The normal distribution is a continuous probability distribution that is symmetric and bell-shaped. Many natural phenomena follow this distribution, such as heights or weights of people. It is fully characterized by its mean ($\mu$) and standard deviation ($\sigma$).

3. **Bernoulli Distribution**: The Bernoulli distribution is a discrete probability distribution that describes a single trial with two possible outcomes, usually labeled as success (with probability $p$) and failure (with probability $1-p$). It is commonly used for modeling binary events.

4. **Binomial Distribution**: The binomial distribution describes the probability of obtaining a specific number of successes in a fixed number of independent Bernoulli trials, where each trial has two possible outcomes (usually referred to as success and failure). It is characterized by two parameters: the number of trials ($n$) and the probability of success ($p$) in each trial.

5. **Poisson Distribution**: The Poisson distribution models the number of events occurring in a fixed interval of time or space. It is often used to model rare events, such as the number of emails received per hour or the number of accidents at a particular intersection. It is characterized by a single parameter $\lambda$, which represents the average rate of occurrence.


### Distributions in R

To see what distributions R supports, use

```
?Distributions
```
or 
```
help(Distributions)
```
for more information. Every distribution has four different functions associated to them. The letter at the beginning of the distribution indicates the functions capabilities.

|Letter|Functionality|
|:-|:-|
|"d"|returns the height of the probability density function|
|"p"|returns the cummulative density function value|
|"q"|returns the inverse cummulative density function (percentiles) |
|"r"|returns a randomly generated number|


### Probabilities

R can compute the probabilities of a distribution given the correct parameters. If you need the cummulative probability, `p` in front of the distribution function is needed. If you need the probability for a discrete distribution, the `d` in front of the distribution function is needed. Below are a few examples.

Find $P(X \leq 5 )$ where $X \sim N(6,2)$. Here we will use the `pnorm()` and we set `q = 5`, `mean = 6` and `sd = sqrt(2)`.

```{r xnorm5, exercise = TRUE}

```

```{r xnorm5-solution}
pnorm(5,6,sqrt(2))
```

Find $P(X \geq 7 )$ where $X \sim N(6,2)$. Here we will use the `pnorm()` and we set `q = 5`, `mean = 6`, `sd = sqrt(2)` and `lower.tail = F`.

```{r xnorm7, exercise = TRUE}

```

```{r xnorm7-solution}
pnorm(7,6,sqrt(2),F)
```

Find $P(X = 20 )$ where $X \sim Bin(30,0.8)$. Here we will use the `dbinom()` and we set `x = 20`, `size = 30` and `prob = 0.8`.

```{r xbinom30, exercise = TRUE}

```

```{r xbinom30-solution}
dbinom(20,30,.8)
```

### Percentiles

Finding the values (percentiles) for a Z, F, and t distributions, use `qnorm()`, `qf()`, and `qt()` functions, respectively.

Finding the $95^{th}$ percentile from $N(0,1)$, we will use the `qnorm()`. We will set `mean=0`, `sd=1`, and `p=0.95`.

```{r z95, exercise = TRUE}

```

```{r z95-solution}
qnorm(.95,0,1)
```


Finding the $95^{th}$ percentile from a t-distribution with DF=9, we will use the `qt()`. We will set `df=9` and `p=0.95`.

```{r t95, exercise = TRUE}

```

```{r t95-solution}
qt(.95,9)
```


Finding the $95^{th}$ percentile for $F(3,4)$, we will use the `qt()`. We will set `df1=3`, `df2=4` and `p=0.95`.

```{r f95, exercise = TRUE}
qf(.95,3,4)
```

```{r f95-solution, exercise = TRUE}

```

### Random Number Generator


R is capable of generating random numbers. For example if we want to generate a random sample of size fifty from a normal distribution with mean eight and variance three, we will use the `rnorm()`. If we want to generate a random sample from any distribution, use the distribution function with `r` in front of it. 

Let's first generate the random sample of fifty from $X \sim N(8,3)$. This is done with the `rnorm()` and setting `n = 50`, `mean = 8` and `sd = sqrt(3)`.

```{r rsnorm, exercise = TRUE}

```

```{r rsnorm-solution}
rnorm(50,8,sqrt(3))
```

No let's generate a random sample of 100 form an `X \sim \Gamma (2,3)`. This is done by using the `rgamma()` and setting `n = 100`, `shape = 2`, and `rate = 3`.

```{r rsgamma, exercise = TRUE}

```


```{r rsgamma-solution}
rgamma(100,2,3)
```


## Statistics

Here, we will go over some of the basic syntax to obtain basic statistics. We will use the variables `mpg` and `cyl` from the `mtcars` data set. To view the data set use the `head()`:

```{r mtcarsview, exercise = TRUE}

```

```{r mtcarsview-solution}
head(mtcars)
```

The variable `mpg` would be used as a continuous variable, and the variable `cyl` would be used as a categorical variable.

###

### Point Estimates

```{r PE, echo=FALSE}
question_text(
  "Describe in your words what is the definition of a point estimate. Click the 'Continue' button to reveal the answer.",
  answer(NULL, correct = TRUE),
  correct = NULL,
  rows = 4,
  try_again_button = "Edit Answer",
  incorrect = NULL,
  allow_retry = TRUE
)

```

###

The first basic statistic you can compute are point estimates. These are your means, medians, etc. Here we will calculate these estimates. 

#### Mean

To obtain the mean, use the `mean()`, you only need to specify `x=` for the data to compute the mean:

```{r meanmpg, exercise = TRUE}

```


```{r meanmpg-solution}
mean(mtcars$mpg)
```

```{r meanmpg-check}
grade_this_code()
```


#### Median

To obtain the median, use the `median()`, you only need to specify `x=` for the data to compute the median:

```{r medianmpg, exercise = TRUE}

```


```{r medianmpg-solution}
median(mtcars$mpg)
```

```{r medianmpg-check}
grade_this_code()
```


#### Frequency

To obtain a frequency table, use the `table()`, you only need to specify the data as the first argument to compute the frequency table:

```{r freqcyl, exercise = TRUE}

```


```{r freqcyl-solution}
table(mtcars$cyl)
```

```{r freqcyl-check}
grade_this_code()
```


#### Proportion

To obtain a the proportions for the frequency table, use the `prop.table()`. However the first argument must be the results from the `table()`. Use the `table()` inside the `prop.table()` to get the proportions:

```{r propcyl, exercise = TRUE}

```


```{r propcyl-solution}
prop.table(table(mtcars$cyl))
```

```{r propcyl-check}
grade_this_code()
```


### Variability

```{r VAR, echo=FALSE}
question_text(
  "Describe in your words what is the definition of variability. Click the 'Continue' button to reveal the answer.",
  answer(NULL, correct = TRUE),
  correct = NULL,
  rows = 4,
  try_again_button = "Edit Answer",
  incorrect = NULL,
  allow_retry = TRUE
)

```


###

Variability statistics can describes the average deviation of data points from a central tendency (point estimate). It can be thought as how far the data points are from the center. R is capable of calculating several variability statistics.

#### Variance

To obtain the variance, use the `var()`, you only need to specify `x=` for the data to compute the variance:

```{r varmpg, exercise = TRUE}

```


```{r varmpg-solution}
var(mtcars$mpg)
```

```{r varmpg-check}
grade_this_code()
```


#### Standard deviation

To obtain the standard deviation, use the `sd()`, you only need to specify `x=` for the data to compute the standard deviation:

```{r sdmpg, exercise = TRUE}

```


```{r sdmpg-solution}
sd(mtcars$mpg)
```

```{r sdmpg-check}
grade_this_code()
```


#### Max and Min

To obtain the max and min, use the `max()` and `min()`, respectively. You only need to specify the data as the first argument to compute the max and min:

```{r minimaxmpg, exercise = TRUE}

```


```{r minimaxmpg-solution}
max(mtcars$mpg)
min(mtcars$mpg)
```

```{r minimaxmpg-check}
grade_this_code()
```


#### Q1 and Q3

To obtain the Q1 and Q3, use the `quantile()` and specify the desired quantile with `probs=`. You only need to specify the data as the first argument and `probs=` (as a decimal) to compute the Q1 and Q3:

```{r Q13mpg, exercise = TRUE}

```


```{r Q13mpg-solution}
quantile(mtcars$mpg,.25)
quantile(mtcars$mpg,.75)
```

```{r Q13mpg-check}
grade_this_code()
```


### Associations

In statistics, we may be interested on how different variables are related to each other. These associations can be represented in a numerical value.

#### Continuous and Continuous

When we measure the association between to continuous variables, we tend to use a correlation statistic. This statistic tells us how linearly associated are the variables are to each other. Essentially, as one variable increases, what happens to the other variable? Does it increase (positive association) or does it decrease (negative association). To find the correlation in R, use the `cor()`. You will need to specify the `x=` and `y=` which represents vectors for each variable. Find the correlation between `mpg` and `hp` from the `mtcars` data set.

```{r corr, exercise = TRUE}

```


```{r corr-solution}
cor(mtcars$mpg,mtcars$hp)
```

```{r corr-check}
grade_this_code()
```


```{r corr-q, echo=FALSE}
question("What type of association does R display?",
  answer("Negative", correct = TRUE),
  answer("No Association"),
  answer("Positive"),
  incorrect = "Incorrect: Check the sign of the number.")

```


#### Categorical and Continuous

When comparing categorical variables, it becomes a bit more nuanced in how to report associations. Most of time you will discuss key differences in certain groups. Here, we will talk about how to get the means for different groups of data. Our continuous variable is the `mpg` variable, and our categorical variable is the `cyl` variable. Both are from the `mtcars` data set.  The `tapply()` allows us to split the data into different groups and then calculate different statistics. We only need to specify `X=` of the R object to split, `INDEX=` which is a list of factors or categories indicating how to split the data set, and `FUN=` which is the function that needs to be computed. Use the `tapply()` and find the mean `mpg` for each `cyl` group: 4, 5, and 6.

```{r cvc, exercise = TRUE}

```


```{r cvc-solution}
tapply(mtcars$mpg, list(mtcars$cyl), mean)
```

```{r cvc-check}
grade_this_code()
```


```{r cvc-q, echo=FALSE}
question("Which cylinder has the smallest mpg?",
  answer("8", correct = TRUE),
  answer("4"),
  answer("6")
)
```



#### Categorical and Categorical

Reporting the association between two categorical variables is may be challenging. If you have a $2\times 2$ table, you can report a ratio of association. However, any other case may be challenging. You can report a hypothesis test to indicate an association, but it does not provide much information about the effect of each variable. You can also report row, column, or table proportions. Here we will talk about creating cross tables and report these proportions. To create a cross table, use the `table()` and use the first two arguments to specify the two categorical variables. Create a cross tabulation between `cyl` and `carb` from the `mtcars` data set.

```{r ctab-00, exercise = TRUE}

```

```{r ctab-00-solution}
table(mtcars$cyl,mtcars$carb)
```

```{r ctab-00-check}
grade_this_code()
```


Notice how the first argument is represented in the rows and the second argument is in the columns. Now create table proportions using both of the variables. You first need to create the table and store it in a variable and then use the `prop.table()`.

```{r ctab_1, exercise = TRUE}

```

```{r ctab_1-solution}
prop.table(table(mtcars$cyl,mtcars$carb))
```

```{r ctab_1-check}
grade_this_code()
```


To get the row proportions, use the argument `margin = 1` within the `prop.table()`.


```{r ctab_2, exercise = TRUE}

```

```{r ctab_2-solution}
prop.table(table(mtcars$cyl,mtcars$carb), margin = 1)
```

```{r ctab_2-check}
grade_this_code()
```


To get the column proportions, use the argument `margin = 2` within the `prop.table()`.


```{r ctab_3, exercise = TRUE}

```

```{r ctab_3-solution}
prop.table(table(mtcars$cyl,mtcars$carb), margin = 2)
```

```{r ctab_3-check}
grade_this_code()
```



## Quiz

```{r quiz, echo=FALSE}
quiz(caption = "",
  question("What is the code to generate 100 random variables from a Bernoulli distribution with parameter p?",
    answer("dbinom(100, 1, p)", message = "The function is mispecified."),
    answer("dbinom(1, 100, p)", message = sprintf("The arguments and function are mispecified.")),
    answer("rbinom(100, 1, p)", correct = TRUE),
    answer("rbinom(1, 100, p)", message = "The arguments are mispecified."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("What are common central tendency statistics?",
    answer("Mean", correct = TRUE),
    answer("Range", message = "Measures lowest and highest values."),
    answer("Median", correct = TRUE),
    answer("Variance.", message = "Measures variability."),
    incorrect = "Incorrect. There is at least on one more thing.",
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
    question("Which 2 statistics can be used to measure the association between two variables?",
    answer("Spearman's Correlation", correct = TRUE),
    answer("Kappa Statistic", message = "Used for interrator reliability."),
    answer("Bayes' Factor", message = "Used for hypothesis testing."),
    answer("Kendall's Tau", correct = TRUE),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("Which functions can be used to obtain statistics on categorical variables?",
    answer("table()", correct = TRUE),
    answer("lm()"),
    answer("prop.table()", correct = TRUE),
    answer("quantile()"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("Use the help(Distributions) command to learn more about R's capability on probability distributions.",
           answer("TRUE", correct = T),
           answer("FALSE"),
           allow_retry = TRUE),
  question("Which letter gives you the quantile of a distribution function",
           answer("d"),
           answer("q"),
           answer("r", correct = TRUE),
           allow_retry = TRUE,
           random_answer_order = TRUE
           ),
  question_text(
  "Write a reflection on what you had learned about basic probability and statistics in R.",
  answer(NULL, correct = TRUE),
  trim = TRUE,
  rows = 4,
  correct = NULL,
  incorrect = NULL,
  allow_retry = TRUE)

  
)
```

## Submit Responses

**Generate a Hash if you need to submit your work for an assignment.**

```{r context="server"}
learnrhash::encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please click the button below to generate your hash and",
    "submit it using the corresponding tutorial assignment tab on your Learning Management System, if needed.",
    shiny::tags$br()
  ))
```

## Feedback

```{=html}
<iframe src="https://docs.google.com/forms/d/e/1FAIpQLSda_bV4VFGk1pgrevik7afgjAlWTA0g9kQE7ZR52S_UQiT3NA/viewform?embedded=true" width="640" height="742" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>
```
