---
title: "R Essentials"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE,
  echo = TRUE
)
```

```{r setup}
library(csustats)
```


Here is a list of tutorials that provide the programming essentials to begin analyzing data in R.

## Basics

```{r}
access_tutorial("R_Basics")
```

This tutorial introduces the basics of R programming.

-   Calculator

-   Operators

-   Functions

-   Data Types

-   Objects

## R Objects

```{r}
access_tutorial("R_Obj")
```

This tutorial introduces users to the basic idea of R objects. It extends topics from the R Basics tutorial.

-   Types of Data

-   R Objects

-   Indexing

-   Loading Data

## Probability and Statistics

```{r}
access_tutorial("R_Stats")
```

This tutorial provides basic code for computing probability and statistics.

## Tidyverse

```{r}
access_tutorial("R_Tidyverse")1
```

This tutorial introduces `dplyr` and `tidyr` packages to users.

-   Pipes

-   Data Transformation

-   Reshaping Data

## ggplot2

```{r}
access_tutorial("R_ggplot2")
```

This tutorial introduces users to the basic idea of the `ggplot2` R package.

-   Basics

-   Grouping

-   Themes

## Strategies

```{r}
access_tutorial("R_Strategy")
```

This tutorial provides tips and tricks to extend your R programming skills in RStudio.

-   RStudio

-   Customizations

-   R Projects

-   R Packages

-   Scripting

-   Debugging
