---
title: "Introduction to R"
output: 
  learnr::tutorial:
    css: css/styles.scss
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: |
  Provides a basic introduction to R Programming.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
tutorial_options(exercise.timelimit = 60, 
                 exercise.checker = gradethis::grade_learnr)

gradethis_setup(exercise.reveal_solution = TRUE)
knitr::opts_chunk$set(echo = FALSE)
```



## Introduction to R {data-progressive="FALSE"}

R is a multi-paradigm programming language that is popular among statisticians. The motto of R is "Created by statisticians, for statisticians". It has many statistical and graphical tools to help statisticians analyze data. Additionally, the programming language can be extended to implement new statistical techniques, create websites, and create dashboards (interactive software). Lastly, and most importantly, R can be downloaded and used for free.

**This tutorial will cover the basics of using R.**

```{r intro, echo=FALSE}
question_text(
  "Make a prediction on what you think R can and cannot do.",
  answer(NULL, correct = TRUE),
  trim = TRUE,
  rows = 4,
  correct = NULL,
  try_again_button = "Edit Answer",
  incorrect = NULL,
  allow_retry = TRUE)

```

## R Calculator

R can be thought as an advanced calculator. Type the following code in the code box and see what R prints out. Does the output correspond to what you expect to see?

Try `9+9`

```{r r-calc-1, exercise = TRUE}

```

```{r r-calc-1-check}
grade_result(
  pass_if(~ (.result == 9 + 9))
)
```

Try `9*9`

```{r r-calc-2, exercise = TRUE}

```

```{r r-calc-2-check}
grade_result(
  pass_if(~ (.result == 9 * 9))
)
```

Try `9-9`

```{r r-calc-3, exercise = TRUE}

```

```{r r-calc-3-check}
grade_result(
  pass_if(~ (.result == 9 - 9))
)
```

Try `9/9`

```{r r-calc-4, exercise = TRUE}

```

```{r r-calc-4-check}
grade_result(
  pass_if(~ (.result == 9 / 9))
)
```

Try `9+9+9`

```{r r-calc-5, exercise = TRUE}

```

```{r r-calc-5-check}
grade_result(
  pass_if(~ (.result == 9 + 9 + 9))
)
```

Try `9+9*9`

```{r r-calc-6, exercise = TRUE}

```

```{r r-calc-6-check}
grade_result(
  pass_if(~ (.result == 9 + 9 * 9))
)
```

Try `9+9/1`

```{r r-calc-7, exercise = TRUE}

```

```{r r-calc-7-check}
grade_result(
  pass_if(~ (.result == 9 + 9 / 1))
)
```

Try `9+(9+9)/9`

```{r r-calc-8, exercise = TRUE}

```

```{r r-calc-8-check}
grade_result(
  pass_if(~ (.result == 9 + (9 + 9) / 2))
)
```

Try `log(9, base=3)`

```{r r-calc-9, exercise = TRUE}

```

```{r r-calc-9-check}
grade_result(
  pass_if(~ (.result == log(9, 3)))
)
```

Try `exp(5)` (e\^5)

```{r r-calc-10, exercise = TRUE}

```

```{r r-calc-10-check}
grade_result(
  pass_if(~ (.result == exp(5)))
)
```
Try `3**3`

```{r r-calc-11, exercise = TRUE}

```

```{r r-calc-11-check}
grade_result(
  pass_if(~ (.result == 3**3))
)
```

Try `pi`

```{r r-calc-12, exercise = TRUE}

```

```{r r-calc-12-check}
grade_result(
  pass_if(~ (.result == pi))
)
```

Try `sin(pi*2)`

```{r r-calc-13, exercise = TRUE}

```

```{r r-calc-13-check}
grade_result(
  pass_if(~ (.result == sin(pi * 2)))
)
```

Try `cos(pi*2)`

```{r r-calc-14, exercise = TRUE}

```

```{r r-calc-14-check}
grade_result(
  pass_if(~ (.result == cos(pi * 2)))
)
```

Try `tanpi(2)`

```{r r-calc-15, exercise = TRUE}

```

```{r r-calc-15-check}
grade_result(
  pass_if(~ (.result == tanpi(2)))
)
```

## R Operators

Operators are symbols that tells R to to conduct an operation. In the previous section, you were tasked to compute the values of certain formulas. To conduct these tasks, you used special symbols (known as operators) that tells R what to do with the numbers.

There are several operators that you can use in R. Generally speaking, they can be classified as arithmetic, comparison, and special operators. Click "Continue" to learn more about them.

###

### Arithmetic Operators

Arithmetic operators are symbols that tells R to conduct an arithmetic operation such as add two numbers together. The table below provides some commonly used operators in R for arithmetic operations:

| Operator | Description        |
|----------|--------------------|
| \+       | Addition           |
| \-       | Subtraction        |
| \*       | Multiplication     |
| /        | Division           |
| \^       | Exponentiation     |
| %%       | Modulo (remainder) |
| %/%      | Integer division   |


### Comparison Operators

Comparison operators will compare to numbers (or data values) and return a logical value (TRUE or FALSE) if the operation is valid or not. The table below provides some commonly used operators in R for comparison operations:

| Operator | Description              |
|----------|--------------------------|
| ==       | Equal to                 |
| !=       | Not equal to             |
| \>       | Greater than             |
| \<       | Less than                |
| \>=      | Greater than or equal to |
| \<=      | Less than or equal to    |
| !        | Negation                 |
| &        | Logical AND              |
| \|       | Logical OR               |


Try `5 <= 2`

```{r r-opp-1, exercise = TRUE}

```

```{r r-opp-1-check}
grade_result(
  pass_if(~ (.result == (5 <= 2)))
)
```

Try `cospi(2) == cos(2* pi)`

```{r r-opp-2, exercise = TRUE}

```

```{r r-opp-2-check}
grade_result(
  pass_if(~ (.result == (cospi(2) == cos(2* pi))))
)
```

Try `76 > 56 | 76 < 56`

```{r r-opp-3, exercise = TRUE}

```

```{r r-opp-3-check}
grade_result(
  pass_if(~ (.result == (76 > 56 | 76 < 56)))
)
```

Try `76 > 56 & 76 < 56`

```{r r-opp-4, exercise = TRUE}

```

```{r r-opp-4-check}
grade_result(
  pass_if(~ (.result == (76 > 56 & 76 < 56)))
)
```


```{r opp-q-1, echo=FALSE}
question_text(
  "Looking at the previous two examples, why do you think you get different results when you switch from '|' to '&'?",
  answer(NULL, correct = TRUE),
  trim = TRUE,
  rows = 4,
  correct = NULL,
  try_again_button = "Edit Answer",
  incorrect = NULL,
  allow_retry = TRUE)

```

### Special Operators

The table below provides some commonly used operators in R for special operations:

| Operator | Description              |
|----------|--------------------------|
| %in%     | Element-wise membership  |
| \<-      | Assignment               |
| -\>      | Right-to-left assignment |
| %\*%     | Matrix multiplication    |
| %o%      | Outer product            |
| %x%      | Kronecker product        |
| \|\>     | Pipe operator            |
| %\>%     | Pipe operator            |

## R Functions

An R function is a procedure that tells R to do certain tasks. For example, the `log(x)` is an R function. It takes the value `x` and provides you the natural logarithm. Here `x` is known as an argument which needs to be specified to us the `log()` function. Find the `log(x = 5)`

```{r r-fx-1, exercise = TRUE}

```

```{r r-fx-1-check}
grade_result(
  pass_if(~ (.result == log(5)))
)
```

Another argument for the `log()` function is the `base` argument. With the previous code, we did not specify the `base` argument, so R makes the `base` argument The natural logarithm is just log with base **e**. If you want to use the common log with base 10, you will need to set the `base` argument equal to 10.

Try `log(x = 5, base = 10)`

```{r r-fx-2, exercise = TRUE}

```

```{r r-fx-2-check}
grade_result(
  pass_if(~ (.result == log(5, 10)))
)
```

Now try `log(5, 10)`

```{r r-fx-3, exercise = TRUE}

```

```{r r-fx-3-check}
grade_result(
  pass_if(~ (.result == log(5, 10)))
)
```

Notice that it provides the same value. This is because R can set arguments based on the position they are provided in the function when it is not specified. For `log(5, 10)`, R thinks that 5 corresponds to the first argument `x` and 10 is the second argument `base`.

## R Data Types

In R, there are several data types available used to conduct complex analysis. These data types tell R how to conduct an analysis. For example, when you give R a number, it will evaluate what type it is and the conduct a task, such as an arithmetic operation. Click "Continue" to see the most commonly used data types.


###


| Data Type | Description                                         |
|-----------|-----------------------------------------------------|
| Numeric   | Represents numeric values (real or decimal numbers) |
| Integer   | Represents whole numbers without decimal points     |
| Logical   | Represents boolean values (TRUE or FALSE)           |
| Character | Represents text or string values                    |
| POSIXct   | Represents date and time values                     |
| NULL      | Represents absence of a value                       |

###

In R, you can test whether a value is a certain data by using the general `is.X()` function where you replace `X` with a data type you suspect.

Try `is.numeric(pi)`

```{r r-fx-4, exercise = TRUE}

```

```{r r-fx-4-check}
grade_result(
  pass_if(~ (.result == is.numeric(pi)))
)
```


Try `is.numeric(TRUE)`

```{r r-fx-5, exercise = TRUE}

```

```{r r-fx-5-check}
grade_result(
  pass_if(~ (.result == is.numeric(TRUE)))
)
```

Try `is.integer(2)`

```{r r-fx-6, exercise = TRUE}

```

```{r r-fx-6-check}
grade_result(
  pass_if(~ (.result == is.integer(2)))
)
```

Try `is.integer(2L)`

```{r r-fx-7, exercise = TRUE}

```

```{r r-fx-7-check}
grade_result(
  pass_if(~ (.result == is.integer(2L)))
)
```


```{r opp-q-2, echo=FALSE}
question_text(
  "Looking at the previous two examples, why do you think you get different results when you and 'L'?",
  answer(NULL, correct = TRUE),
  trim = TRUE,
  rows = 4,
  correct = NULL,
  try_again_button = "Edit Answer",
  incorrect = NULL,
  allow_retry = TRUE)

```


## R Objects

One major component in R is R objects. R objects are can be thought of as virtual containers of things. These objects can store results, numbers, functions, and many more. Here were will work on a few concepts with R objects. First, let's ;earn how to create an R object. To create an R object called `x` that contains the number 12, use the `<-` operator. The `<-` operator is just the less than arrow `<` and minus sign `-`. Type `x <- 12`.

```{r r-obj-1-setup}
x <- 12
```

```{r r-obj-1, exercise = TRUE}

```

```{r r-obj-1-solution}
x <- 12
```

```{r r-obj-1-check}
grade_result(
  pass_if(~ identical(.result, x <- 12))
)
```

Notice that nothing pops out. This is because R does not need to print out any results. Now to view what is in `x`, just type `x`.

```{r r-obj-2, exercise = TRUE, exercise.setup = "r-obj-1-setup"}

```

```{r r-obj-2-solution}
# Arrow ----
x

# Non Arrow ----
print(x)
```

```{r r-obj-2-check}
grade_this_code()
```

Notice that the number 12 pops out.

Now try to add something to the R object `x`. Type `x + 3`

```{r r-obj-3, exercise = TRUE, exercise.setup = "r-obj-1-setup"}

```

```{r r-obj-3-solution}
 x + 3
```

```{r r-obj-3-check}
grade_this_code()
```

Notice that 15 prints out.

Now try other calculations to `x`

```{r r-obj-4, exercise = TRUE, exercise.setup = "r-obj-1-setup"}

```

R objects are powerful and versatile. More information can be found the R Objects tutorial which can be accessed here:

```{default}
csustats::access_tutorial("R_Obj")
```

## Quiz

```{r quiz, echo=FALSE}
quiz(caption = "",
  question("Which function do you need to use to properly load an R package?",
    answer("install.packages()", message = "This will install an R package."),
    answer("load()", message = sprintf("This is used to load data, not a package.")),
    answer("library()", correct = TRUE),
    answer("require()", message = "While this will work, it is advised not to use this function."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("What is the code to add 5 to 4 squared?",
    answer("5 + 4**2", correct = TRUE),
    answer("5 * 4 && 4", message = "Shiny apps are used to create data dashboards"),
    answer("4^2 + 5", correct = TRUE),
    answer("4*4 + 5", correct = TRUE),
    incorrect = "Incorrect. There is at least one more thing.",
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
    question("Which operator is used to create an R object with data? ",
    answer("<-", correct = TRUE),
    answer("==", message = "This is the absolute equal operator."),
    answer("=", correct = TRUE),
    answer("<<", message = "This is not a known operator"),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("Which one is the pipe operator?",
    answer("|>", correct = TRUE),
    answer("%|%", message = "This is not a known R operator."),
    answer("%>%", correct = TRUE),
    answer("%%", message = "This is the modulo (remainder) operator."),
    allow_retry = TRUE,
    random_answer_order = TRUE
  ),
  question("Do you need to specify the name of the argument for a function?",
           answer("TRUE"),
           answer("FALSE", correct = T),
           allow_retry = TRUE),
  question("Which function tells you if a data value is a Logical value?",
           answer("if.logical()"),
           answer("as.logical()"),
           answer("is.logical()", correct = TRUE),
           allow_retry = TRUE,
           random_answer_order = TRUE
           ),
  question_text(
  "Write a reflection on what you had learned about R Basics.",
  answer(NULL, correct = TRUE),
  trim = TRUE,
  rows = 4,
  try_again_button = "Edit Answer",
  correct = NULL,
  incorrect = NULL,
  allow_retry = TRUE)

  
)
```

## Submit Responses

**Generate a Hash if you need to submit your work for an assignment.**

```{r context="server"}
learnrhash::encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please click the button below to generate your hash and",
    "submit it using the corresponding tutorial assignment tab on your Learning Management System, if needed.",
    shiny::tags$br()
  ))
```

## Feedback

```{=html}
<iframe src="https://docs.google.com/forms/d/e/1FAIpQLSda_bV4VFGk1pgrevik7afgjAlWTA0g9kQE7ZR52S_UQiT3NA/viewform?embedded=true" width="640" height="742" frameborder="0" marginheight="0" marginwidth="0">Loading…</iframe>
```
