---
title: "One- and Two-Sample Hypothesis Testing"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)

y3 <- rnorm(25, 5, 1)
y4 <- y3 - rnorm(2, .25, 1)

pharm <- data.frame(pre = y3, post = y4)

diff <- pharm$post - pharm$pre
```

## Introduction

This tutorial focuses on t-test hypothesis testing and checking assumptions. At the end of the tutorial, there are problems about conducting t-tests. Since this tutorial focuses on the t-test, we will mainly be using the `t.test()`. This function contains all the functionality you need to conduct a t-test in R.

Through out this tutorial, we use certain notations for different components in R. To begin, when something is in a gray block, `_`, this indicates that R code is being used. When I am talking about an R Object, it will be displayed as a word. For example, we will be using the R object `mtcars`. When I am talking about an R function, it will be displayed as a word followed by an open and close parentheses. For example, we will use the mean function denoted as `mean()` (read this as "mean function"). When I am talking about an R argument for a function, it will be displayed as a word following by an equal sign. For example, we will use the data argument denoted as `data=` (read this as "data argument"). When I am referencing an R package, I will use `::` (two colons) after the name. For example, in this tutorial, I will use the `ggplot2::` (read this as "ggplot2 package") Lastly, if I am displaying R code for your reference or to run, it will be displayed on its own line. There are many components in R, and my hope is that this will help you understand what components am I talking about.

### Contents

1.  1-Sample t-test

2.  2-Sample Independent t-test

3.  Paired t-test

## 1-Sample t-test

A 1-sample t-test is used when you want to test the mean of a sample to population mean. Here we will give a small review on the assumptions of a t-test and R code to test the assumptions and conduct the test. For this section, we will be using the `mpg` variable from the `mtcars` data set.

Before we begin, find the mean, variance and sample size for `mpg`.

```{r tt1_1_mean, exercise = TRUE}

```

```{r tt1_1_mean-solution}
mean(mtcars$mpg)
var(mtcars$mpg)
length(mtcars$mpg)
```

Notice how the sample size is greater than 30. This means we can use a Z-test to conduct the hypothesis test. However, a t-test becomes closer to a Z-distribution when the sample size gets larger. At 30, the t-distribution is close to the Z-distribution. Therefore, we can use R's t-test which will give similar results to a Z-test.

### Assumptions

Before we can begin testing the sample mean with the population mean, we must check the assumptions.

#### Assumptions

```{r tt1_1, echo = FALSE}
question("What are the assumptions for a 1-sample t-test? (select ALL that apply)",
  answer("Data is from a normal population", correct = TRUE),
  answer("Observations are continuous", correct = TRUE),
  answer("Observations are independent of each other", correct = TRUE),
  answer("The true variance is unknown", correct = TRUE),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

As the question above indicates there are several assumptions that we must check. First let's check the normality assumption.

```{r tt1_2, echo = FALSE}

question("What methods do you use to test the data comes from a normal distribution? (select ALL that apply)",
  answer("Shapiro-Wilks Test", correct = TRUE),
  answer("Levene Test"),
  answer("QQ-Plot", correct = TRUE),
  answer("F-Test"),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

#### Normality

##### Shapiro-Wilk Test

The Shapiro-Wilk Test is another hypothesis test. It tests the following hypotheses:

$H_0:$ The sample comes from a normal distribution.

$H_1:$ The sample does not come from a normal distribution.

Therefore, we look at the p-value and conduct the hypothesis test.

In R, we conduct a Shapiro-Wilk Test using the `shapiro.test()`. All you need to specify is a vector of data.

```{r tt1_3, exercise = TRUE}

```

```{r tt1_3-solution}
shapiro.test(mtcars$mpg)
```

```{r tt1_4, echo = FALSE}
question("Does the data follow a normal distribution?",
 answer("TRUE", correct = TRUE),
 answer("FALSE"),
 incorrect = NULL
)
```

##### QQ-Plot

The QQ-plot is a graphical tool used to assess the whether the data follows a normal distribution or not. If the data follows a normal distribution, it will have a qq-plot like below:

```{r tt1_5, echo = FALSE}
x_norm <- rnorm(10000)
qqnorm(x_norm)
qqline(x_norm)
```

Notice that all the dots follow along the line. This is what we want. All the points need to be as close as the line as possible. Any deviations indicate a violation of the assumption as shown below.

```{r tt1_6, echo = FALSE}
x_cauchy <- runif(10000)
qqnorm(x_cauchy)
qqline(x_cauchy)
```

The tails deviate much from the line which indicates a violation of the assumption.

To create a QQ Plot in R, we will use the `qqnorm()` and the `qqline()`. The `qqnorm()` function will create the QQ Plot. The `qqline()` adds the guiding line to assess the assumption. Both functions only require a vector of data. Additionally, in order to make both of these functions work, they must run at the same time. So highlight both to run them. Otherwise, R will output an error of a missing graphic. Check the assumption on the variable `mpg` in `mtcars`.

```{r tt1_7, exercise = TRUE}

```

```{r tt1_7-solution}
qqnorm(mtcars$mpg)
qqline(mtcars$mpg)
```

#### Independence

There are statistical tests for this assumption, but for right not we will assume they are independent from each other.

#### Continuous

The definitions of continuous is a bit tricky to understand and confirm. However, to run a t-test, the main thing is that the data is a number. That means it can be discrete or continuous.

#### Outliers

The best way to check if there are outliers is by plotting either the histogram or box plot. Plot the histogram of the data.

```{r tt1_8, exercise = TRUE}

```

```{r tt1_8-solution}
hist(mtcars$mpg)
```

```{r tt1_9, echo = FALSE}
question("Are there any potential outliers?",
  answer("TRUE", correct = TRUE, message = "However, for the purpose of this tutorial we will assume that there are no outliers."),
  answer("FALSE"),
  incorrect = "The data is right skewed. However, for the purpose of this tutorial we will assume that there are no outliers."
 )
```

For now, we will assume there are no outliers.

### t-test

As stated before, to conduct a t-test in R, we will use the `t.test()`. This function contains all the functionality to 1-sample and 2-sample tests as well as 1-sided and 2-sided tests. Everything you learn here is applicable for the other types of t-tests.

Before we begin, to specify the data needed to conduct the test, we use the `x=`. This is the first argument in the function. The data just needs to be as a vector.

#### One-Sided Test

A one-sided test occurs when the alternative hypothesis is contains a less than ($<$) or greater than ($>$). The `t.test()` accounts for the one-sided test using the `alternative=`.

##### Less than test

Test the following hypotheses:

$H_0: \mu = 25$

$H_1: \mu \lt 25$

Since the population mean is 25, we will need to set the `mu=` to 25. Additionally, we will set the `alternative=` equal to `"less"`. Now, conduct the test.

```{r tt1_10, exercise = TRUE}

```

```{r tt1_10-solution}
t.test(mtcars$mpg, mu = 25, alternative = "less")
```

```{r tt1_11, echo = FALSE}
question("Do you reject the null hypothesis at an $\\alpha=0.05$?",
  answer("Yes", correct = TRUE, message = "The p-value is less than $\\alpha$."),
  answer("No"),
  incorrect = "The p-value is less than $\\alpha$."
 )
```

##### Greater than test

Test the following hypotheses:

$H_0: \mu = 20$

$H_1: \mu \gt 20$

We need to do is set the `alternative=` equal to `"greater"` and set the `mu=` to 20. Now, conduct the test.

```{r tt1_12, exercise = TRUE}

```

```{r tt1_12-solution}
t.test(mtcars$mpg, mu = 20, alternative = "greater")
```

```{r tt1_13, echo = FALSE}
question("Do you reject the null hypothesis at an $\\alpha=0.05$?",
  answer("Yes"),
  answer("No", correct = TRUE, message = "The p-value is greater than $\\alpha$."),
  incorrect = "The p-value is greater than $\\alpha$."
 )
```

##### 

#### Two-sided Test

Test the following hypotheses:

$H_0: \mu = 20$

$H_1: \mu \neq 20$

We need to do is set the `alternative=` equal to `"two.sided"` and set the `mu=` to 20. Now, conduct the test.

```{r tt1_14, exercise = TRUE}

```

```{r tt1_14-solution}
t.test(mtcars$mpg, mu = 20, alternative = "two.sided")
```

```{r tt1_15, echo = FALSE}
question("Do you reject the null hypothesis at an $\\alpha=0.05$?",
          answer("Yes"),
          answer("No", correct = TRUE, message = "The p-value is greater than $\\alpha$."),
  incorrect = "The p-value is greater than $\\alpha$."
 )
```

##### 

## 2-Sample Independent t-test

For most research, we are comparing populations to further our understanding. To compare different population means, we use the 2-sample independent t-test. This t-test is different from the paired t-test both mathematically and fundamentally. A paired t-test indicates that the data from the 2 samples are linked to each other in some way. The 2-sample independent t-test does not have the linkage. Here we will talk about how to conduct a 2-sample independent t-test. For the rest of this section, we will compare the means of the mean mpg (variable `mpg`) for two different types of engines (variable `vs`, only two categories labeled as 0 or 1) in the `mtcars` data set.

To begin, find the mean, variance, and sample size for each group (Note: the `tidyverse::` is automatically loaded in this tutorial if you want to use that instead).

```{r tt2_1, exercise = TRUE}

```

```{r tt2_1-solution}
tapply(mtcars$mpg, list(mtcars$vs), mean)
tapply(mtcars$mpg, list(mtcars$vs), var)
tapply(mtcars$mpg, list(mtcars$vs), length)

#OR

mtcars %>% group_by(vs) %>%
  summarise(mean = mean(mpg), var = var(mpg), size = length(mpg))
```

### Assumptions

As with the 1-sample t-test, the 2-sample t-test has assumptions that need to be satisfied to conduct the test properly.

#### Assumptions

```{r tt2_2, echo = FALSE}
question("What are the assumptions for a 2-sample independent t-test? (select ALL that apply)",
  answer("Data is from a normal population", correct = TRUE),
  answer("Observations are continuous", correct = TRUE),
  answer("Observations are independent of each other", correct = TRUE),
  answer("There are no outliers", correct = TRUE),
  answer("The variance for both populations are the same", correct = TRUE),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

As you can see, many assumptions are the same as a 1-sample t-test. The only new assumption is that the variance is the same for each population. If this assumption is not satisfied, a different t-test must be used. R's `t.test()` is capable of doing this as well.

#### Normality

Test the assumption of that the populations come from a normal distribution, you will need to test it for each group. Test the assumption for each group. You may need to create vectors for each group.

```{r tt2_3, exercise = TRUE}

```

```{r tt2_3-solution}
tapply(mtcars$mpg, list(mtcars$vs), shapiro.test)

mpg_vs0 <- mtcars$mpg[mtcars$vs == 0]
mpg_vs1 <- mtcars$mpg[mtcars$vs == 1]

qqnorm(mpg_vs0)
qqline(mpg_vs0)

qqnorm(mpg_vs1)
qqline(mpg_vs1)


```

There may be a violation due to the qq plot. The data may need to be transformed to ensure that the data is normal. For now, we will assume the assumption is satisfied.

#### Equal Variance

To test for the equal variance assumption, we can use the `var.test()` which conducts an F-test with the following assumptions:

$H_0: \sigma_1=\sigma_2$

$H_1: \sigma_1 \neq \sigma_2$

The `var.test()` requires two arguments, `x=` and `y=`, to specify the vectors of data. Test the assumption.

```{r tt2_4, exercise = TRUE}

```

```{r tt2_4-solution}
mpg_vs0 <- mtcars$mpg[mtcars$vs == 0]
mpg_vs1 <- mtcars$mpg[mtcars$vs == 1]

var.test(mpg_vs0, mpg_vs1)
```

```{r tt2_5, echo = FALSE}
question("Do you reject the null hypothesis at an $\\alpha=0.05$?",
          answer("Yes"),
          answer("No", correct = TRUE, message = "The p-value is greater than $\\alpha$."),
  incorrect = "The p-value is greater than $\\alpha$."
 )
```

The equal variance assumption is satisfied. Let's us assume all the other assumptions are satisfied.

### t-test

Here we will show you how to conduct a 2-sample t-test for both situations for the equal variance assumption. Before we talk about the different situations, the input for the data in the `t.test()` can be either two separate vectors (`x=` and `y=`) or as a formula and specify the data set (`formula=y~x`, first argument, and the `data=`). If you choose to use the formula method, your `x` variable must specify which observation belongs to which group.

For the rest of this section, we will test the following.

$H_0: \mu_1 = \mu_2$

$H_1: \mu_1 \neq \mu_2$

The default argument the `alternative=` is `"two.sided"`; therefore, you do not need to specify it. Additionally, the default argument for the `mu=` is `0`.

#### Equal variance

```{r tt2_6, echo = FALSE}
question("When we have equal variance, what do we need to calculate?",
          answer("Satterwaitte DF"),
          answer("Pooled Variance", correct = TRUE),
  incorrect = NULL
 )
```

To conduct this t-test, you will need to set the `var.equal=` to `TRUE`. Then R will conduct the t-test and use the pooled variance.

```{r tt2_7, exercise = TRUE}

```

```{r tt2_7-solution}
t.test(mpg~vs, data = mtcars, var.equal = TRUE)
```

#### Unequal variance

```{r tt2_8, echo = FALSE}
question("When we have equal variance, what do we need to calculate?",
          answer("Satterwaitte DF"),
          answer("Pooled Variance", correct = TRUE),
  incorrect = NULL
 )
```

To conduct this t-test, you will need to set the `var.equal=` to `FALSE`. However, the default setting for the `var.equal` is `FALSE`. Then R will conduct the t-test and use the separate variances and calculate the Satterwaitte DF.

```{r tt2_9, exercise = TRUE}

```

```{r tt2_9-solution}
t.test(mpg~vs, data = mtcars)
```

## Paired t-test

To me, the trickiest part is selecting the independent t-test vs the paired t-test, What I usually do is try to prove that the samples are linked. If I can't do that, I will then assume it is independent samples. The only quick trick is if there are different sample sizes, then it is independent t-test. However, equal sample sizes does not mean the paired t-test should be used.

When deciding between independent vs paired, I try to see if one group is linked to another group. I focus on finding the link. If I can't find the link, then I will assume they are independent samples.

Anyways, for the rest of this section, we will use the `pharm` data set to see if the variable `pre` is larger after treatment with the variable `post` variable. We will use the two vectors approach for the `t.test()`.

When conducting a paired t-test, we need to take the difference of the two variables. Find the mean and variance of the difference in the values.

```{r tt3_1, exercise = TRUE}

```

```{r tt3_1-solution}
diff <- pharm$pre - pharm$post
mean(diff)
var(diff)

# OR

pharm %>% mutate(diff = pre - post) %>%
  summarise(mean(diff),var(diff))
```

### Assumptions

The assumptions for the paired t-test are similar to the 1-sample t-test. However, we focus on the difference in the data instead.

#### Assumptions

```{r tt3_2, echo = FALSE}
question("What are the assumptions for a paired t-test? (select ALL that apply)",
  answer("The difference of data follows a normal population", correct = TRUE),
  answer("Observations are continuous", correct = TRUE),
  answer("Observations are independent of each other", correct = TRUE),
  answer("There are no outliers", correct = TRUE),
  random_answer_order = TRUE,
  allow_retry = TRUE
)
```

#### Normality

Test the normality of the difference in data.

```{r tt3_3, exercise = TRUE}

```

```{r tt3_3-solution}
diff <- pharm$post - pharm$pre

shapiro.test(diff)

qqnorm(diff)
qqline(diff)
```

### t-test

To conduct a t-test in R, set the `paired=` equal to `TRUE` in the `t.test()`.

Test the following hypotheses:

$H_0: D = 0$

$H_1: D > 0$

where

$D=pre-post.$

Note: in the `t.test()`, a paired t-test is conducted as $D=x-y$, corresponding to the arguments `x=` (first argument) and `y=` (second argument).

```{r tt3_4, exercise = TRUE}

```

```{r tt3_4-solution}
t.test(pharm$pre, pharm$post, paired = T)
```

## Questions

```{r ttq_1, echo=FALSE}
question("What type of distribution is used to test the mean when the sample size is less than 30 and sigma is unknown?",
         answer("Z", correct = TRUE),
         answer("t"),
         answer("F"),
         answer("$\\chi ^ 2$"),
         allow_retry = TRUE,
         random_answer_order = TRUE
  
)
```

```{r ttq_2, echo=FALSE}
question("What are the degrees of freedom when n = 25?",
         answer("26", correct = TRUE),
         answer("25"),
         answer("24"),
         answer("23"),
         allow_retry = TRUE,
         random_answer_order = TRUE,
         incorrect = "DF = n-1"
)
```

```{r ttq_3, echo=FALSE}
quiz(caption = "Paired Data: Select if the scenario is paired data or not.",
  question("Participants are asked to apply for a credit card at both a credit union and traditional. Their interest rates for each credit card are recorded for further analysis",
         answer("TRUE", correct = TRUE),
         answer("FALSE"),
         incorrect = "Because it is the same participant applying for both credit cards, it links the data together."
  ),
    question("The Sacramento River runs through downtown Sacramento. Water samples were collected both upstream and downstream of the city for analysis.",
         answer("TRUE", correct = TRUE),
         answer("FALSE"),
         incorrect = "Because the river goes through the city, it links data together."
  ),
    question("Researchers wanted to know how access affects a patient's outcome for a disease. An equal number of patients were recruited from a community and private hospital.",
         answer("TRUE"),
         answer("FALSE", correct = TRUE),
         incorrect = "Even though each group has the same number of participants, there is nothing linking a patient from one type of hospital to another."
  )
)
```




## Submit

```{r context="server"}
learnrhash::encoder_logic()
```

```{r encode, echo=FALSE}
learnrhash::encoder_ui(
  ui_before = shiny::div(
    "If you have completed this tutorial and are happy with all of your",
    "solutions, please click the button below to generate your hash and",
    "submit it using the corresponding tutorial assignment tab on your Learning Management System, if needed.",
    shiny::tags$br()
  ))
```
